---
# title: "exercise data wrangling"
# author: "Tianyi Wu"
# date: "2023-09-30"
#output: html_document
#editor_options: 
  #chunk_output_type: console
editor_options: 
  chunk_output_type: inline
---

# Housekeeping

```{r}
library("tidyverse")
d <- read.csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```

# **Codebook lookup**:

## Indicators regarding the quality of education

Education 15+ (E) (e_peaveduc)

Educational inequality, Gini (E) (e_peedgini)

## What are the data's coverage (i.e., for which countries and years do we have data?)

e_peaveduc: 1820-2022

e_peedgini: 1850-2010

## What are their sources? Provide the link to least 1 source.

# **Subset by columns**

## Create a dataset containing only the country-year identifiers and indicators of education quality.

```{r}
d_edu <- d |> select(country_name, year, e_peaveduc, e_peedgini)
```

Rename the columns of education quality to make them informative.

```{r}
d_edu <- d_edu |> rename ("Country"="country_name", "Year"="year", "Edu15P"="e_peaveduc", "EduIne"="e_peedgini")
```

# **Subset by rows**

List 5 countries-years that have the highest education level among its population.

```{r}
d_edu |>
  slice_max(order_by = Edu15P, n = 5)
```

List 5 countries-years that suffer from the most severe inequality in education.

```{r}
d_edu |>
  slice_max (order_by = EduIne, n=5)
```

# **Summarize the data**

Check data availability: For which countries and years are the indicators of education quality available?

```{r}

```

Create two types of country-level indicators of education quality
