---
# title: "exercise data wrangling"
# author: "Tianyi Wu"
# date: "2023-09-30"
#output: html_document
#editor_options: 
  #chunk_output_type: console
editor_options: 
  chunk_output_type: inline
---

# Housekeeping

```{r}
library("tidyverse")
d <- read.csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```

# **Codebook lookup**:

## Indicators regarding the quality of education

Education 15+ (E) (e_peaveduc)

Educational inequality, Gini (E) (e_peedgini)

## What are the data's coverage (i.e., for which countries and years do we have data?)

e_peaveduc: 1820-2022

e_peedgini: 1850-2010

## What are their sources? Provide the link to least 1 source.

e_peaveduc:

<https://clio-infra.eu/Indicators/AverageYearsofEducation.html>

e_peedgini:

<https://clio-infra.eu/Indicators/EducationalInequalityGiniCoefficient.html>

# **Subset by columns**

## Create a dataset containing only the country-year identifiers and indicators of education quality.

```{r}
d_edu <- d |> select(country_name, year, e_peaveduc, e_peedgini)
```

Rename the columns of education quality to make them informative.

```{r}
d_edu <- d_edu |> rename ("Country"="country_name", "Year"="year", "Edu15P"="e_peaveduc", "EduIne"="e_peedgini")
```

# **Subset by rows**

List 5 countries-years that have the highest education level among its population.

```{r}
d_edu |>
  slice_max(order_by = Edu15P, n = 5)
```

List 5 countries-years that suffer from the most severe inequality in education.

```{r}
d_edu |>
  slice_max (order_by = EduIne, n=5)
```

# **Summarize the data**

## Check data availability

## For which countries and years are the indicators of education quality available?

```{r}
# For which countries are the indicators of education quality available
d_edu |>
  mutate(Edu15P_missing = as.numeric(is.na(Edu15P)), EduIne_missing = as.numeric(is.na(EduIne))) |>
  group_by(Country)|>
  summarise(N_Edu15P_missing = sum(Edu15P_missing), N_EduIne_missing = sum(EduIne_missing))
```

```{r}
# For which years are the indicators of education quality available
d_edu |>
  mutate(Edu15P_missing = as.numeric(is.na(Edu15P)), .after= Edu15P, EduIne_missing = as.numeric(is.na(EduIne))) |>
  group_by(Year)|>
  summarise(N_Edu15P_missing = sum(Edu15P_missing), N_EduIne_missing = sum(EduIne_missing))
```

## Create two types of country-level indicators of education quality

```         
1.  Average level of education quality from 1984 to 2022

2.  Change of education quality from 1984 to 2022
```

```{r}
# Calculate the average level of education quality
# Edu15P: from 1984-2022
# EduIne: from 1984-2010 (Seen the available time range in codebook)
d_edu |>
  group_by(Country)|>
  summarise(Edu15P_average = mean (Edu15P, na.rm = TRUE), EduIne_average = mean (EduIne, na.rm = TRUE))
```

```{r}
# Calculate the change of education quality (overall)
# Edu15P: from 1984-2022
# EduIne: from 1984-2010 (Seen the available time range in codebook)
d_edu |>
  group_by(Country) |>
  arrange(Year) |>
  summarise(Change_Edu15P = last(Edu15P) - first(Edu15P))

d_edu |>
  filter (Year >= 1984 & Year <=2010) |>
  group_by(Country) |>
  arrange(Year) |>
  summarise(Change_EduIne = last(EduIne) - first(EduIne))
```

```{r}
# Calculate the change of education quality (year-over-year)
d_edu |>
  group_by(Country) |>
  arrange(Year) |>
  mutate (Edu15P_yoy_change = Edu15P - lag(Edu15P, n=1), .after = Edu15P) |>
  mutate (EduIne_yoy_change = EduIne - lag(EduIne, n=1), .after = EduIne) |>
  ungroup() |>
  arrange(Country, Year)
```

## Examine the data and *briefly* discuss

Which countries perform the best and the worst in terms of education quality in the past four decades?

```{r}
d_edu |>
  group_by(Country) |>
  summarise (Edu15P_mean = mean (Edu15P)) |>
  arrange(Edu15P_mean)
```

Burkina Faso performs the worst while Germany performs the best in terms of education level 15+.

```{r}
d_edu |>
  filter (Year >= 1984 & Year <= 2010)|>
  group_by(Country) |>
  summarise (EduIne_mean = mean (EduIne)) |>
  arrange(EduIne_mean, na.rm = TRUE)
```

Burkina Faso performs the worst while Austria performs the best in terms of education inequality.
